Parallel Simulation of Adaptive Random Boolean Networks

### 2016-07-14 Paper/paper.tex
Restructure the materials in Experimental Study, add speedup description

	\begin{algorithm}[ht!]
		\label{algo:gpu}
		GenerateNetwork()\;
		$epoch \leftarrow 0$\;
		\While{ $epoch < epochsCount$ }
		{
			$gpuNetwork \leftarrow$ ConvertNetworkToGpuRepresentation()\;
			ZeroInitialize($stateSum$)\;
			CopyToGpu($gpuNetwork$, $currentState$, $stateSum$)\;
			\While{attractor not found}
			{ 	
				UpdateStateOnGpu($currentState$, $stateSum$)\;
			}
			CopyFromGpu($stateSum$)\;
			$nodesToRewire \leftarrow$ ChooseRandomNodes($N_{rew}$)\;
			ActivityDependentRewiring($nodesToRewire$, $stateSum$)\;
			$epoch \leftarrow epoch + 1$\;
		}
		\caption{General algorithm of network evolution on GPU}
	\end{algorithm}

	\paragraph{} The performance of the GPU implementation highly depends on the Boolean network representation it uses. Since a node with inputs \(\mathbf{N_i}\) has \(2^{|\mathbf{N_i}|}\) possible combinations of inputs' state, it is a common approach to store a Boolean function of the node as \(1 \times 2^{|\mathbf{N_i}|}\) vector \(\mathbf{b_i}\). We can represent these vectors as \(N \times 2^{\max{|\mathbf{N_i}|}}\) matrix \(B\).

@@

	To test the applicability of the approaches presented in Section \ref{sect:algorithm}, a set of experiments was performed. The experiments were conducted on a cluster consisting of 20 nodes with NVidia GeForce GT 640 graphic cards. The results below were obtained by averaging over 10 realizations of the evolution process.

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/evolution}
			\caption{Evolution of mean network connectivity}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/n_rew}
			\caption{Evolution of mean connectivity for different number of nodes rewired each epoch; $N=1000$}
		\end{minipage}
	\end{figure}

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\textwidth]{plots/attractors}
			\caption{Empirical distribution of attractor lengths}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/speedup}
			\caption{Execution time of the GPU algorithm compared to the serial one}
		\end{minipage}
	\end{figure}

	\begin{table}[ht!]
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	&	750\\ \hline
			$5 \cdot 10^4$	&	99.77	&	88.19	&	71.19	&	45.48	&	29.62\\
			$2.5 \cdot 10^5$	&	99.99	&	95.92	&	81.24	&	56.10	&	35.96\\
			$5 \cdot 10^5$	&	100.00	&	97.43	&	86.58	&	59.36	&	38.73\\
			$2.5 \cdot 10^6$	&	100.00	&	99.14	&	92.99	&	68.74	&	46.46\\
			$5 \cdot 10^6$	&	100.00	&	99.53	&	94.40	&	73.00	&	49.22\\
			$1 \cdot 10^7$	&	100.00	&	99.75	&	96.08	&	77.73	&	53.81\\
		\end{tabular}
		\caption{Percentage $R$ of attractors found during the evolution process with attractor length limeted by $\mu_{max}$}
	\end{table}
	\begin{table}[ht!]
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	& 750\\
			\hline
			$5 \cdot 10^4$	&	2.64	&	2.59	&	2.58	&	2.52	&	2.47\\
			$2.5 \cdot 10^5$	&	2.66	&	2.60	&	2.56	&	2.51	&	2.48\\
			$5 \cdot 10^5$	&	2.68	&	2.59	&	2.55	&	2.51	&	2.48\\
			$2.5 \cdot 10^6$	&	2.69	&	2.60	&	2.55	&	2.51	&	2.48\\
			$5 \cdot 10^6$	&	2.69	&	2.59	&	2.55	&	2.52	&	2.48\\
			$1 \cdot 10^7$	&	2.68	&	2.62	&	2.55	&	2.51	&	2.48\\
		\end{tabular}
		\caption{Mean steady state connectivity $K_{ss}$ obtained as a result of evolution process with attractor length limeted by $\mu_{max}$}
	\end{table}

>>>>

	\begin{algorithm}[ht!]
		\label{algo:gpu}
		GenerateNetwork()\;
		$epoch \leftarrow 0$\;
		\While{ $epoch < epochsCount$ }
		{
			$gpuNetwork \leftarrow$ ConvertNetworkToGpuRepresentation()\;
			ZeroInitialize($stateSum$)\;
			CopyToGpu($gpuNetwork$, $currentState$, $stateSum$)\;
			\While{attractor not found}
			{ 	
				UpdateStateOnGpu($currentState$, $stateSum$)\;
			}
			CopyFromGpu($stateSum$)\;
			$nodesToRewire \leftarrow$ ChooseRandomNodes($N_{rew}$)\;
			ActivityDependentRewiring($nodesToRewire$, $stateSum$)\;
			$epoch \leftarrow epoch + 1$\;
		}
		\caption{General algorithm of network evolution with GPU acceleration}
	\end{algorithm}

	The performance of the GPU implementation highly depends on the Boolean network representation it uses. Since a node with inputs \(\mathbf{N_i}\) has \(2^{|\mathbf{N_i}|}\) possible combinations of inputs' state, it is a common approach to store a Boolean function of the node as \(1 \times 2^{|\mathbf{N_i}|}\) vector \(\mathbf{b_i}\). We can represent these vectors as \(N \times 2^{\max{|\mathbf{N_i}|}}\) matrix \(B\).

@@

	To test the applicability of the approaches presented in Section \ref{sect:algorithm}, a set of experiments was performed. The experiments were conducted on a cluster consisting of 20 nodes with NVidia GeForce GT 640 graphic cards. In this study we considered only the networks with initial in-degree connectivity $K_{ini}=2$. The results were averaged over 10 realizations of the evolution process.

	\paragraph{Speedup obtained by GPU acceleration.} We measured the execution times of both serial and parallel versions of state updates. We used Liu-Bassler algorithm with checkpoints $T$ = $\{100,$ $200,$ $1000,$ $2000,$ $...,$ $10^7,$ $2 \cdot 10^7\}$, $\mu_{max} = 10^7$. To decrease the number of random samples required to compare the versions, we did not stop the epoch as soon as an attractor was found. Instead, we performed all $2 \cdot 10^7$ state updates per epoch. Such simulation corresponds to the worst-case scenario when no attractors were found, i. e. $R = 0\%$. The results presented in Figure \ref{fig:speedup} indicate that the GPU implementation is more efficient than the serial one for networks consisting of more than $200$ nodes.

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/speedup}
			\caption{Execution time of the GPU algorithm compared to the serial one}
			\label{fig:speedup}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/n_rew}
			\caption{Evolution of mean connectivity for different number of nodes rewired each epoch; $N=1000$}
		\end{minipage}
	\end{figure}

	\paragraph{Rewiring more than one node per epoch.}

	\paragraph{Limiting the maximum attractor length.}

	\begin{table}[ht!]
		\centering
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	&	750\\ \hline
			$5 \cdot 10^4$	&	99.77	&	88.19	&	71.19	&	45.48	&	29.62\\
			$2.5 \cdot 10^5$	&	99.99	&	95.92	&	81.24	&	56.10	&	35.96\\
			$5 \cdot 10^5$	&	100.00	&	97.43	&	86.58	&	59.36	&	38.73\\
			$2.5 \cdot 10^6$	&	100.00	&	99.14	&	92.99	&	68.74	&	46.46\\
			$5 \cdot 10^6$	&	100.00	&	99.53	&	94.40	&	73.00	&	49.22\\
			$1 \cdot 10^7$	&	100.00	&	99.75	&	96.08	&	77.73	&	53.81\\
		\end{tabular}
		\caption{Percentage $R$ of attractors found during the evolution process with attractor length limited by $\mu_{max}$}
	\end{table}
	\begin{table}[ht!]
		\centering
		\begin{tabular}{r | c c c c c}
			& \multicolumn{5}{c}{Nodes count} \\
			$\mu_{max}$	& 100	& 	200	&	300	&	500	& 750\\
			\hline
			$5 \cdot 10^4$	&	2.64	&	2.59	&	2.58	&	2.52	&	2.47\\
			$2.5 \cdot 10^5$	&	2.66	&	2.60	&	2.56	&	2.51	&	2.48\\
			$5 \cdot 10^5$	&	2.68	&	2.59	&	2.55	&	2.51	&	2.48\\
			$2.5 \cdot 10^6$	&	2.69	&	2.60	&	2.55	&	2.51	&	2.48\\
			$5 \cdot 10^6$	&	2.69	&	2.59	&	2.55	&	2.52	&	2.48\\
			$1 \cdot 10^7$	&	2.68	&	2.62	&	2.55	&	2.51	&	2.48\\
		\end{tabular}
		\caption{Mean steady state connectivity $K_{ss}$ obtained as a result of evolution process with attractor length limited by $\mu_{max}$}
	\end{table}

	\begin{figure}[ht!]
		\begin{minipage}[t]{0.45\textwidth}
			\includegraphics[width=1.0\textwidth]{plots/evolution}
			\caption{Evolution of mean network connectivity}
		\end{minipage}\hfill
		\begin{minipage}[t]{0.45\textwidth}
			\centering
			\includegraphics[width=1.0\textwidth]{plots/attractors}
			\caption{Empirical distribution of attractor lengths}
		\end{minipage}
	\end{figure}
